---
title: "Consolidation Model Functions"
author: "Gabriel Watson"
date: "2026-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Libraries
```{r}
library(aws.s3)
library(tidyverse)
library(sf)
library(stringr)
#library(purr)
library(tidyr)
library(osrm)
```


## Data Import 
```{r}
national_water_system <- s3read_using(readRDS,
                                      object = "s3://tech-team-data/national-dw-tool/clean/national/national_water_system.RData")

pwsid_huc <- s3read_using(read.csv, object = "s3://tech-team-data/national-dw-tool/test-staged/pwsid_npdes_usts_rmps_imp.csv")%>%
             select(pwsid, num_facilities)%>%
             group_by(pwsid)%>%
             summarize(num_facilities = sum(num_facilities, na.rm = TRUE))

epa_sabs <- national_water_system[[1]] %>% st_as_sf()%>%
            left_join(.,pwsid_huc)

#join_check <- epa_sabs %>% filter(is.na(num_facilities))
```


## Helper Functions ##

## 1. get_sabs
## Takes a vector of SAB ids OR state abv and County Name
## 'small' returns just pwsid,pws_name, and states_intersect
```{r}
get_sabs <- function(ids,size){
  
if(size == "small")
{
sabs <- epa_sabs %>%
        filter(pwsid %in% ids)%>%
        select(pwsid,pws_name,epic_states_intersect, service_connections_count, population_served_count, num_facilities)
}
else
{
sabs <- epa_sabs %>%
        filter(pwsid %in% ids)
}
  
return(sabs)
}
```




```{r}
library(openrouteservice)

# Set your ORS API key once at the top of your script
ors_api_key("eyJvcmciOiI1YjNjZTM1OTc4NTExMTAwMDFjZjYyNDgiLCJpZCI6IjkxMDRlNmE2YmY0YzRhNmE4NGM4NTlmMTNiNmFjZWNjIiwiaCI6Im11cm11cjY0In0=")

# Helper to replicate osrmTable using ORS matrix endpoint
ors_distance_matrix <- function(src_geom, dst_geoms) {
  # Build coordinate list: [longitude, latitude]
  src_coords  <- st_coordinates(st_centroid(src_geom))
  dst_coords  <- st_coordinates(st_centroid(dst_geoms))
  
  all_coords <- rbind(src_coords, dst_coords) %>%
    as.data.frame() %>%
    transmute(lon = X, lat = Y) %>%
    pmap(~ c(..1, ..2))  # ORS expects list of c(lon, lat)
  
  n_src <- nrow(src_coords)
  n_dst <- nrow(dst_coords)
  
  # ORS matrix: sources = index 0, destinations = indices 1..n
  result <- ors_matrix(
    locations  = all_coords,
    sources    = 0:(n_src - 1),
    destinations = n_src:(n_src + n_dst - 1),
    metrics    = "distance",
    profile    = "driving-car"
  )
  
  # Returns distances in metres — same shape as osrmTable$distances
  list(distances = matrix(result$distances, nrow = n_src))
}
```


## 2. get_neighbors
## Takes a df of pwsids
## calcs centroid
## calcs crow-flies distance between points
## calcs distance as transit distance
## pwsids, list(pwsid,centroid_distance,travel_distance)
## TO DO - include all systems that just overlap no matter the centroid distance
## TO DO - only run pairs on defined filter config for receiving and consolidating systems (see other TO DO in the logic run)
```{r}
get_neighbors <- function(sabs, cutoff) {
  message("Calculating centroids...")
  centroids <- st_centroid(sabs$geometry)
  centroid_distance_matrix <- st_distance(centroids, centroids)
  
  message("Processing ", nrow(sabs), " pwsids...")
  
  df <- sabs %>%
    mutate(rec = map(1:n(), function(i) {
      if (i %% 10 == 0) message("  Processing pwsid ", i, " of ", n())
      
      centroid_distances <- as.numeric(centroid_distance_matrix[i, ]) * 0.000621371
      matches <- tibble(
        pwsid            = pwsid[-i],
        centroid_distance = centroid_distances[-i]
      ) %>%
        filter(centroid_distance <= cutoff)
      
      if (nrow(matches) == 0) {
        return(tibble(pwsid = character(), centroid_distance = numeric(),
                      travel_distance = numeric(), overlap = logical()))
      }
      
      parent_geom <- sabs$geometry[i]
      child_geoms <- sabs$geometry[sabs$pwsid %in% matches$pwsid]
      
      # --- Replaced osrmTable with ORS ---
      travel_table <- tryCatch(
        ors_distance_matrix(parent_geom, child_geoms),
        error = function(e) {
          message("  ORS error for pwsid ", i, ": ", e$message,
                  " — falling back to centroid distance")
          NULL
        }
      )
      
      matches %>%
        mutate(
          travel_distance = if (!is.null(travel_table)) {
            as.numeric(travel_table$distances) * 0.000621371
          } else {
            centroid_distance   # graceful fallback
          },
          overlap = st_overlaps(parent_geom, child_geoms, sparse = FALSE)[1, ]
        ) %>%
        filter(travel_distance <= cutoff)
    })) %>%
    mutate(rec_num_neighbors = map_int(rec, nrow))
  
  message("Unnesting results...")
  df <- unnest(df, rec, names_sep = "_")
  
  df <- df %>%
    left_join(
      sabs %>%
        data.frame() %>%
        select(pwsid, num_facilities) %>%
        rename(rec_pwsid = pwsid, rec_num_facilities = num_facilities)
    ) %>%
    mutate(rec_num_facilities = ifelse(is.na(rec_num_facilities), 0, rec_num_facilities))
  
  message("Complete! Found neighbors for ", nrow(df), " pairs")
  return(df)
}


```



## Takes a neighbor df
## Calcs Costs 

## Additional Input
## xcost per mile 
## xconnnection fees
## xservice line costs 
## xadmin costs


## determined
## xCEQA costs (based on intersect )
## Additional costs 

## Overhead/Additional 
## Contingency
## Planning and construction costs
## Engineering Services
## Inflation
## Regional Multiplier
```{r}
get_costs <- function(neighbors,
                       cost_per_mile,
                       connection_fee,
                       service_line_fee, 
                       admin_cost,
                       contingency_const,
                       planning_constuction_const,
                       engineering_services_const,
                       inflation_const,
                       regional_multiplier_const
                       )
{
  
  df <- neighbors %>%
        mutate(new_source_cost = ifelse(rec_num_facilities > 0,0,1238933))%>% ## new source cost of $1,238,933 based on Greg's calcs 
        mutate(pipe_line_cost = (rec_travel_distance * cost_per_mile))%>%
        mutate(connection_fees = service_connections_count * connection_fee)%>%
        mutate(service_line_cost = service_connections_count *service_line_fee)%>%
        mutate(admin_costs = (pipe_line_cost +service_line_cost) * admin_cost)%>%
        mutate(CEQA_cost = ifelse(rec_overlap == TRUE,25000,100000))%>%
        rowwise()%>%
        mutate(total_capital_costs = sum(c_across(new_source_cost:CEQA_cost)))%>%
        mutate(contingency = total_capital_costs * contingency_const)%>%
        mutate(planning_constuction = total_capital_costs * planning_constuction_const)%>%
        mutate(engineering_services = total_capital_costs * engineering_services_const)%>%
        mutate(inflation = total_capital_costs * inflation_const)%>%
        mutate(regional_multiplier = total_capital_costs * regional_multiplier_const)%>%
        rowwise()%>%
        mutate(total_markup = sum(c_across(contingency:regional_multiplier)))%>%
        mutate(total_project_cost = total_markup + total_capital_costs)
  
}

```


## filter_sabs
## Consolidation Parms 
## cons_system
## - tbd 
## - Violations
## - ownership
## - Service connection count
## rec_system
## - tbd 
## - Violations 
## - ownership
## - Service connection count
```{r}
filter_sabs <- function(sabs, cons_sys_config, rec_sys_config)
{
  
sdwis_sabs <- national_water_system[[2]] %>%
              select(pwsid,owner_type,health_viols_10yr,open_health_viol)
  
owner_type_con <- cons_sys_config %>% pull(owner_type)
owner_type_rec <-rec_sys_config %>% pull(owner_type)

print(owner_type_con)
cons_sabs <- sabs %>% 
             left_join(.,sdwis_sabs)%>%
             filter(health_viols_10yr >= cons_sys_config %>% pull(health_viols_10yr))%>%
             filter(open_health_viol == open_health_viol)%>%
             filter(owner_type == owner_type_con)%>%
             filter(population_served_count <= cons_sys_config %>% pull(pop_served))%>%
             pull(pwsid)

rec_sabs <- sabs %>% 
             left_join(.,sdwis_sabs)%>%
             filter(health_viols_10yr <= rec_sys_config %>% pull(health_viols_10yr))%>%
             filter(open_health_viol == open_health_viol)%>%
             filter(owner_type == owner_type_rec)%>%
             filter(population_served_count >= rec_sys_config %>% pull(pop_served))%>%
             pull(pwsid)

sabs_list <- list()
sabs_list[[1]] <- cons_sabs 
sabs_list[[2]] <- rec_sabs

return(sabs_list)
}

``` 

## regional stats
```{r}
get_regional_costs <- function(costs)
{
  
regional_costs  <- costs %>%
                   data.frame()%>%
                   select(-c(geometry))%>%
                   arrange(pwsid,total_project_cost)%>%
                   distinct(pwsid, .keep_all = TRUE)%>%
                   summarise(across(c(new_source_cost:total_project_cost), sum))
                  
  
}


```


## Testing Corner
```{r}

## 1. Select based on state ##
## TO DO: Select by county (need county match - not in data) ## 
sab_test <- get_sabs(epa_sabs %>% filter(str_detect(epic_states_intersect, "MD")) %>% pull(pwsid),"small")


### FILTER SABS #### x
## quick mvp config for consolidating systems and receiving systems
## TO DO ## Add all option to select all ownership types
cons_sys_config <- data.frame(
                  open_health_viol= c("No"),
                  health_viols_10yr = c(1),
                  owner_type = c("Private"),
                  pop_served = c(1000))

rec_sys_config <- data.frame(
                  open_health_viol = c("No"),
                  health_viols_10yr = c(1),
                  owner_type = c("Local"),
                  pop_served = c(10000))

## returns list of sabs viable for receiving and consolidation 
filtered_sabs_config <- filter_sabs(sab_test,cons_sys_config,rec_sys_config)

## runs filter on state universe
sab_test_filtered <- sab_test %>%
                    filter(pwsid %in% filtered_sabs_config[[1]] | pwsid %in% filtered_sabs_config[[2]])

## calculates the neighbor/consolidation pairs
## TO DO - send the config to the get_neighbors and double check there so only receiving systems can receive and consolidating systems can consolidate - do this before running travel distance pairs so we don't obliterate the API with unneeded requests. 
neighbors <- get_neighbors(sab_test_filtered,5) %>%
             filter(pwsid %in% filtered_sabs_config[[1]])%>%
             filter(rec_pwsid %in% filtered_sabs_config[[2]])



## config for costs ## 
cost_per_mile <- 1000000
connection_fee <- 4000
service_line_fee <- 6200
admin_cost <- .15
contingency_const <- .2
planning_constuction_const <- .1
engineering_services_const <- .15
inflation_const <- .031
regional_multiplier_const <- .1
               

## getting costs ## 
costs <- get_costs(neighbors,
                       cost_per_mile,
                       connection_fee,
                       service_line_fee, 
                       admin_cost,
                       contingency_const,
                       planning_constuction_const,
                       engineering_services_const,
                       inflation_const,
                       regional_multiplier_const)


## getting regional costs ## 
regional_costs <- get_regional_costs(costs) 



```



