---
title: "Consolidation Model Functions"
author: "Gabriel Watson"
date: "2026-01-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r}
library(aws.s3)
library(tidyverse)
library(sf)
library(stringr)
library(tidyr)
library(openrouteservice)
```

## Data Import

```{r}
national_water_system <- s3read_using(
  readRDS,
  object = "s3://tech-team-data/national-dw-tool/clean/national/national_water_system.RData"
)

pwsid_huc <- s3read_using(
  read.csv,
  object = "s3://tech-team-data/national-dw-tool/test-staged/pwsid_npdes_usts_rmps_imp.csv"
) %>%
  select(pwsid, num_facilities) %>%
  group_by(pwsid) %>%
  summarize(num_facilities = sum(num_facilities, na.rm = TRUE))

epa_sabs <- national_water_system[[1]] %>%
  st_as_sf() %>%
  left_join(., pwsid_huc)
```

## Helper Functions

#### 1. get_sabs
Takes a vector of SAB ids OR state abv.
'small' returns just pwsid, pws_name, and states_intersect.

```{r}
get_sabs <- function(ids, size) {
  if (size == "small") {
    sabs <- epa_sabs %>%
      filter(pwsid %in% ids) %>%
      select(pwsid, pws_name, epic_states_intersect,
             service_connections_count, population_served_count, num_facilities)
  } else {
    sabs <- epa_sabs %>%
      filter(pwsid %in% ids)
  }
  return(sabs)
}
```

```{r}
ors_api_key("eyJvcmciOiI1YjNjZTM1OTc4NTExMTAwMDFjZjYyNDgiLCJpZCI6IjkxMDRlNmE2YmY0YzRhNmE4NGM4NTlmMTNiNmFjZWNjIiwiaCI6Im11cm11cjY0In0=")

ors_distance_matrix <- function(src_geom, dst_geoms) {
  src_coords <- st_coordinates(st_centroid(src_geom))
  dst_coords <- st_coordinates(st_centroid(dst_geoms))

  all_coords <- rbind(src_coords, dst_coords) %>%
    as.data.frame() %>%
    transmute(lon = X, lat = Y) %>%
    pmap(~ c(..1, ..2))

  n_src <- nrow(src_coords)
  n_dst <- nrow(dst_coords)

  result <- ors_matrix(
    locations    = all_coords,
    sources      = 0:(n_src - 1),
    destinations = n_src:(n_src + n_dst - 1),
    metrics      = "distance",
    profile      = "driving-car"
  )

  list(distances = matrix(result$distances, nrow = n_src))
}
```

## 2. get_neighbors
#Takes a df of pwsids, cons_pwsids, and rec_pwsids.
#- Calcs centroid and crow-flies distance between points.
#- **Pre-filters to only cons→rec pairs before any ORS call** to avoid burning API quota.
#- **Always includes pairs whose service area boundaries overlap**, regardless of centroid distance.
#- Calcs travel distance only for qualifying pairs.

```{r}
get_neighbors <- function(sabs, cutoff, cons_pwsids, rec_pwsids) {

  message("Calculating centroids...")
  centroids <- st_centroid(sabs$geometry)
  centroid_distance_matrix <- st_distance(centroids, centroids)

  message("Processing ", nrow(sabs), " pwsids...")

  df <- sabs %>%
    mutate(rec = map(1:n(), function(i) {

      if (i %% 10 == 0) message("  Processing pwsid ", i, " of ", n())

      this_pwsid <- pwsid[i]

      # --- Pre-filter: only process consolidating systems as the parent ---
      if (!(this_pwsid %in% cons_pwsids)) {
        return(tibble(
          pwsid             = character(),
          centroid_distance = numeric(),
          travel_distance   = numeric(),
          overlap           = logical()
        ))
      }

      parent_geom <- sabs$geometry[i]

      # --- Pre-filter: candidate children must be receiving systems ---
      candidate_idx <- which(pwsid != this_pwsid & pwsid %in% rec_pwsids)

      if (length(candidate_idx) == 0) {
        return(tibble(
          pwsid             = character(),
          centroid_distance = numeric(),
          travel_distance   = numeric(),
          overlap           = logical()
        ))
      }

      centroid_distances <- as.numeric(centroid_distance_matrix[i, candidate_idx]) * 0.000621371
      candidate_pwsids   <- pwsid[candidate_idx]
      child_geoms        <- sabs$geometry[candidate_idx]

      # --- Overlap check ---
      overlaps <- st_overlaps(parent_geom, child_geoms, sparse = FALSE)[1, ]

      # --- Keep pairs within cutoff OR overlapping before hitting OSRM ---
      matches <- tibble(
        pwsid             = candidate_pwsids,
        centroid_distance = centroid_distances,
        overlap           = overlaps
      ) %>%
        filter(centroid_distance <= cutoff | overlap)

      if (nrow(matches) == 0) {
        return(tibble(
          pwsid             = character(),
          centroid_distance = numeric(),
          travel_distance   = numeric(),
          overlap           = logical()
        ))
      }

      matched_geoms <- sabs$geometry[sabs$pwsid %in% matches$pwsid]

      message("  Finished calculating centroids, running OSRM for pwsid ", i)

      travel_table <- tryCatch(
        osrmTable(
          src     = st_centroid(parent_geom),
          dst     = st_centroid(matched_geoms),
          measure = "distance"
        ),
        error = function(e) {
          message("  OSRM error for pwsid ", i, ": ", e$message,
                  " — falling back to centroid distance")
          NULL
        }
      )

      message("  Finished calculating travel distance for pwsid ", i)

      matches %>%
        mutate(
          travel_distance = if (!is.null(travel_table)) {
            as.numeric(travel_table$distances[1, ]) * 0.000621371
          } else {
            centroid_distance  # graceful fallback
          },
          overlap = st_overlaps(parent_geom, matched_geoms, sparse = FALSE)[1, ]
        ) %>%
        filter(travel_distance <= cutoff | overlap)

    })) %>%
    mutate(rec_num_neighbors = map_int(rec, nrow))

  message("Unnesting results...")
  df <- unnest(df, rec, names_sep = "_")

  df <- df %>%
    left_join(
      sabs %>%
        data.frame() %>%
        select(pwsid, num_facilities) %>%
        rename(rec_pwsid = pwsid, rec_num_facilities = num_facilities)
    ) %>%
    mutate(rec_num_facilities = ifelse(is.na(rec_num_facilities), 0, rec_num_facilities))

  message("Complete! Found neighbors for ", nrow(df), " pairs")
  return(df)
}
```

## 3. get_costs

```{r}
get_costs <- function(neighbors,
                      cost_per_mile,
                      connection_fee,
                      service_line_fee,
                      admin_cost,
                      contingency_const,
                      planning_constuction_const,
                      engineering_services_const,
                      inflation_const,
                      regional_multiplier_const) {

  df <- neighbors %>%
    mutate(new_source_cost   = ifelse(rec_num_facilities > 0, 0, 1238933)) %>%
    mutate(pipe_line_cost    = rec_travel_distance * cost_per_mile) %>%
    mutate(connection_fees   = service_connections_count * connection_fee) %>%
    mutate(service_line_cost = service_connections_count * service_line_fee) %>%
    mutate(admin_costs       = (pipe_line_cost + service_line_cost) * admin_cost) %>%
    mutate(CEQA_cost         = ifelse(rec_overlap == TRUE, 25000, 100000)) %>%
    rowwise() %>%
    mutate(total_capital_costs  = sum(c_across(new_source_cost:CEQA_cost))) %>%
    mutate(contingency          = total_capital_costs * contingency_const) %>%
    mutate(planning_constuction = total_capital_costs * planning_constuction_const) %>%
    mutate(engineering_services = total_capital_costs * engineering_services_const) %>%
    mutate(inflation            = total_capital_costs * inflation_const) %>%
    mutate(regional_multiplier  = total_capital_costs * regional_multiplier_const) %>%
    rowwise() %>%
    mutate(total_markup       = sum(c_across(contingency:regional_multiplier))) %>%
    mutate(total_project_cost = total_markup + total_capital_costs)
}
```

## 4. filter_sabs
Consolidation and receiving system filters.
**owner_type accepts "All" to skip ownership filtering** for either config.

```{r}
filter_sabs <- function(sabs, cons_sys_config, rec_sys_config) {

  sdwis_sabs <- national_water_system[[2]] %>%
    select(pwsid, owner_type, health_viols_10yr, open_health_viol)

  owner_type_con <- cons_sys_config %>% pull(owner_type)
  owner_type_rec <- rec_sys_config  %>% pull(owner_type)

  cons_sabs <- sabs %>%
    left_join(., sdwis_sabs) %>%
    filter(health_viols_10yr >= cons_sys_config %>% pull(health_viols_10yr)) %>%
    filter(open_health_viol  == cons_sys_config %>% pull(open_health_viol)) %>%
    # "All" skips ownership filter
    { if (owner_type_con != "All") filter(., owner_type == owner_type_con) else . } %>%
    filter(population_served_count <= cons_sys_config %>% pull(pop_served)) %>%
    pull(pwsid)

  rec_sabs <- sabs %>%
    left_join(., sdwis_sabs) %>%
    filter(health_viols_10yr <= rec_sys_config %>% pull(health_viols_10yr)) %>%
    filter(open_health_viol  == rec_sys_config  %>% pull(open_health_viol)) %>%
    # "All" skips ownership filter
    { if (owner_type_rec != "All") filter(., owner_type == owner_type_rec) else . } %>%
    filter(population_served_count >= rec_sys_config %>% pull(pop_served)) %>%
    pull(pwsid)

  list(cons = cons_sabs, rec = rec_sabs)
}
```

## 5. get_regional_costs

```{r}
get_regional_costs <- function(costs) {
  costs %>%
    data.frame() %>%
    select(-c(geometry)) %>%
    arrange(pwsid, total_project_cost) %>%
    distinct(pwsid, .keep_all = TRUE) %>%
    summarise(across(c(new_source_cost:total_project_cost), sum))
}
```

## Testing Corner

```{r}
# 1. Select based on state
sab_test <- get_sabs(
  epa_sabs %>% filter(str_detect(epic_states_intersect, "TX")) %>% pull(pwsid),
  "small"
)

# 2. Config for consolidating and receiving systems
#    Set owner_type = "All" to skip ownership filtering
cons_sys_config <- data.frame(
  open_health_viol  = "No",
  health_viols_10yr = 1,
  owner_type        = "Private",   # or "All"
  pop_served        = 1000
)

rec_sys_config <- data.frame(
  open_health_viol  = "No",
  health_viols_10yr = 1,
  owner_type        = "Local",     # or "All"
  pop_served        = 10000
)

# 3. Returns named list: $cons and $rec
filtered_sabs_config <- filter_sabs(sab_test, cons_sys_config, rec_sys_config)

# 4. Restrict universe to systems in either role
sab_test_filtered <- sab_test %>%
  filter(pwsid %in% filtered_sabs_config$cons |
         pwsid %in% filtered_sabs_config$rec)

# 5. get_neighbors now receives the role lists directly —
#    pre-filters to cons→rec pairs before any ORS call,
#    and always includes overlapping pairs regardless of distance.
neighbors <- get_neighbors(
  sabs        = sab_test_filtered,
  cutoff      = 3,
  cons_pwsids = filtered_sabs_config$cons,
  rec_pwsids  = filtered_sabs_config$rec)

# 6. Cost config
cost_per_mile             <- 1000000
connection_fee            <- 4000
service_line_fee          <- 6200
admin_cost                <- 0.15
contingency_const         <- 0.2
planning_constuction_const <- 0.1
engineering_services_const <- 0.15
inflation_const           <- 0.031
regional_multiplier_const <- 0.1

# 7. Costs
costs <- get_costs(
  neighbors,
  cost_per_mile,
  connection_fee,
  service_line_fee,
  admin_cost,
  contingency_const,
  planning_constuction_const,
  engineering_services_const,
  inflation_const,
  regional_multiplier_const
)

# 8. Regional rollup
regional_costs <- get_regional_costs(costs)

nrow(sab_test_filtered)
length(filtered_sabs_config$cons)
length(filtered_sabs_config$rec)
```